---
title: "code_for_model"
author: "Zongchao Liu"
date: "12/6/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(arsenal)
library(kableExtra)
library(broom)
```



#table 1
## clean data first
```{r,include=FALSE}
data = read_csv("Lawsuit.csv") %>%janitor::clean_names() %>%
  mutate( dept = factor(dept),
        gender = factor(gender),
         clin = factor(clin),
         cert = factor(cert),
         rank = factor(rank)) %>%
  mutate(dept = recode(dept,
                       `1` = "Biochemistry/Molecular Biology",
                       `2` = "Physiology",
                       `3` = "Genetics",
                       `4` = "rankPediatrics",
                       `5` = "Medicine",
                       `6` = "Surgery"),
         gender = recode(gender,
                         `1` = "male",
                         `0` = "female"),
         clin = recode(clin,
                       `1` = "Primarily clinical emphasis",
                       `0` = "Primarily research emphasis"),
         cert = recode(cert,
                       `1` = "Board certified", 
                       `0` = "not certified"),
         rank = recode(rank,
                       `1`= "Assistant", 
                       `2`= "Associate", 
                       `3`= "Full professor")) %>%
  select(-id)

```


## make table 1
```{r}
control <- arsenal::tableby.control(test = F,
                                    numeric.stats = c("meansd","medianq1q3","range"),
                                    ordered.stats = c("Nmiss"),
                                    digits = 2)
tbl <- arsenal::tableby( gender ~ ., data = data, control = control)

summary(tbl,text = T) %>%
  knitr::kable(booktabs = T, caption = "Descriptive Statistics") %>%
  kable_styling(latex_options = c("striped", "hold_position")) 
# remeber to add title!
```

# data cleaning for modeling
## use sal94 as the outcome
```{r}
data = read_csv("Lawsuit.csv") %>%janitor::clean_names() %>%
  mutate( dept = factor(dept),
        gender = factor(gender),
         clin = factor(clin),
         cert = factor(cert),
         rank = factor(rank)) %>%
 # mutate(dept = recode(dept,
 #                      `1` = "Biochemistry/Molecular Biology",
 #                      `2` = "Physiology",
 #                      `3` = "Genetics",
 #                      `4` = "rankPediatrics",
 #                      `5` = "Medicine",
 #                      `6` = "Surgery"),
 #        gender = recode(gender,
 #                        `1` = "male",
 #                        `0` = "female"),
 #        clin = recode(clin,
 #                      `1` = "Primarily clinical emphasis",
 #                      `0` = "Primarily research emphasis"),
 #        cert = recode(cert,
 #                      `1` = "Board certified", 
 #                      `0` = "not certified"),
 #        rank = recode(rank,
 #                      `1`= "Assistant", 
 #                      `2`= "Associate", 
 #                      `3`= "Full professor")) %>%
  select(gender,everything())  %>%
  select(-id, -sal95) %>%
  mutate(sal94 = log(sal94))


```


```{r}
# fit
attach(data)
#based model
fit.base = lm(sal94 ~ gender)
summary(fit.base) #0.38624

# see if there is any confounder
output_model = vector()
output_add_var = vector()

for (i in 2:7 ) {
  var = names(data)[i] # ensure correct var names
  formula = as.formula(str_c("sal94 ~ gender +", var))
  print(str_c("sal94 ~ gender +", var))
  output_model = append(output_model, str_c("sal94 ~ gender +", var))
  output_add_var = append(output_add_var, var)
  }

confounder_check = tibble(model = output_model,
                             add_var = output_add_var)

regression = function(formula){
  formula = as.formula(formula)
  fit = lm(formula , data = data)
  return(as.tibble(broom::tidy(fit)))
}

confounder_check = 
  confounder_check %>%
  mutate(reg = map(confounder_check$model,regression)) %>%
  unnest() %>%
  filter(p.value < .05) %>%
  filter(term == "gender1") %>%
  mutate(confounder = ifelse(abs((estimate - 0.38624)/0.38624) > 0.1, "confounder","not confounder" ))
# rank is not a confounder

#
summary(lm(sal94 ~ gender + dept + clin + cert + prate +exper, data = data))

# all other variables are confounders and should be adjusted. Next, let's test the interaction effect between gender and other factors.


```


```{r}
# test interaction between gender and other varibales
output_model = vector()
output_add_var = vector()

for (i in 2:6 ) {
  var = names(data)[i] # ensure correct var names
  formula = as.formula(str_c("sal94 ~ gender + dept + clin + cert + prate + exper +","gender*", var))
  print(str_c("sal94 ~ gender + dept + clin + cert + prate +exper + gender *", var))
  output_model = append(output_model, str_c("sal94 ~ gender + dept + clin + cert + prate +exper + gender *", var))
  output_add_var = append(output_add_var, var)
}

inter_check = 
  tibble(model = output_model,
       inter_with = output_add_var) %>%
  mutate(reg = map(model,regression)) %>%
  unnest() %>%
  filter(p.value < .05) # clin , exper are interaction terms that should be included


summary(lm(sal94 ~ gender + dept + clin + cert + prate + exper + gender * clin + gender * exper , data = data))

interact.fit_1 = lm(sal94 ~ gender + dept + clin + cert + exper + gender * clin + gender * exper , data = data)
summary(interact.fit_1)
interact.fit_2 = lm(sal94 ~ gender + dept + clin + cert + prate + exper + gender * clin + gender * exper , data = data)
summary(interact.fit_2)

anova(interact.fit_1,interact.fit_2) # anova shows that smaller model(only without prate) is better  (interact.fit_1)
# however, prate is a confounder, we still should include that variable 
#the resulting model is as below:
summary(interact.fit_2)
```

The resulting model is:
$$Ln(sal94)  = 10.99 + \mbox{0.28gender1 -0.143dept2 + 0.19dept3 + 0.15dept4 + 0.49dept5 + 0.21clin1 + 0.185cert1 + 0.04exper - 0.120gender1 * clin1 -0.018 gender1 * exper}$$

# Diagnostic

## leverage values, outliers, influential points

```{r}
par(mfrow = c(2,2))
plot(interact.fit_2)
```

The residuals follow normal distribution and have constant variance. Using Cook's distance, we find that the 184th observation might be an influential point.


```{r}
# outliers in Y
std_resid = rstandard(interact.fit_2)
y_outliers = std_resid[abs(std_resid > 2.5)] # 2 outliers

# outliers in X
influential_points = as.data.frame(influence.measures(interact.fit_2)[[1]]) %>%
  mutate(id = c(1:261)) %>%
  select(id,everything())
nrow(influential_points %>% filter(hat > 0.5)) # 0 leverage values

# influencial points - dffit 
influential_points %>% filter(dffit > 1) #row 184 -> influential point
influential_points %>% filter(dffit > 2*sqrt(12/261)) # 8,73,101,182, 184, 239 influential point

# influencial points - cook's distance
influential_points %>% filter(cook.d > 12/261) # row 184 -> influential point
```
# Colinearity
```{r}
library(corrplot)
library(HH)
data %>%
  select(5,6,8) %>%
  cor() %>%
  corrplot.mixed(upper = "ellipse")
vif(interact.fit_2)
```
`prate` has a high VIF, indicating a high multicolinearity. However, it is used as a control variable of the model. Such a high VIF is not a problem and can be safely ignored. （https://statisticalhorizons.com/multicollinearity）

# stratified models
In the `model_94`,we have 5 confounders: `dept`, `clin`, `cert`, `prate`, `exper`.

```{r}
data_dept1 = data%>%
  filter(dept == "1") 
lm(sal94 ~ gender + dept + clin + cert + prate + exper + gender * clin + gender * exper, data_dept1)
```

## stratified by `dept`, `clin`, `cert`:

```{r}
#dept :1 ~ 6
dept_stratify = function(i){
  data_dept = data %>%
    filter(dept == as.character(i))
  summary(lm(sal94 ~ gender  + clin + cert + prate + exper + gender * clin + gender * exper, data_dept))
}
dept_stratify(1)
dept_stratify(2)
dept_stratify(3)
dept_stratify(4)
dept_stratify(5)
dept_stratify(6)


# cert 0, 1
cert_stratify = function(i){
  data_cert = data %>%
    filter(cert == as.character(i))
  summary(lm(sal94 ~ gender  + dept + clin + prate + exper + gender * clin + gender * exper, data_cert))
}

cert_stratify(0)
cert_stratify(1)

# clin: 0, 1
clin_stratify = function(i){
  data_clin = data %>%
    filter(clin == as.character(i))
  summary(lm(sal94 ~ gender  + dept + cert + prate + exper + gender * exper, data_clin))
}

clin_stratify(0)
clin_stratify(1)

```



